---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# verbalisr

<!-- badges: start -->
<!-- badges: end -->

The purpose of **verbalisr** is to describe pedigree relationships in plain language. This is often helpful in order to understand complex genealogies. Given two members of any pedigree, **verbalisr** spells out the connecting paths between them, using common terminology like *great-grandmother* and *half first cousins*. 

**verbalisr** is part of the **ped suite** framework for pedigree analysis in R.

## Installation

The development version of **verbalisr** can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("magnusdv/verbalisr")
```
## Example

```{r example}
library(verbalisr)
```

Here is an example involving a double-cousin-like relationship:

```{r fig.keep = "none"}
x = doubleCousins(degree1 = 1, removal1 = 1,
                  degree2 = 2, removal2 = 0)

plot(x, hatched = 15:16)
```

```{r dblcous, echo = FALSE, fig.height = 3.5, fig.width = 5, out.width = "60%"}
plot(x, hatched = leaves(x), margin = c(1,1,1,1))
```


To get a written description of the two youngest members, we use `verbalise()`:
```{r}
verbalise(x, ids = 15:16)
```


